*Prochlorococcus marinus* responses to light and oxygen

### 2024-03-07 Code/Manuscript_O2.Rmd
Summary edits and Abstract draft 248 words

@@

We analyzed  growth rates to determine the viability of *P. marinus* MED4, a clade HLI ecotype found near the ocean surface,
*P. marinus* SS120, a clade LLII/III ecotype found deep in the water column, and *P. marinus* MIT9313, a clade LLIV ecotype also found in deep oceans including OMZ, under a matrix of spectral waveband, light level, photoperiod and oxygen conditions approximating present day and hypothetical future niches.

*Prochlorococcus marinus* MED4 requires more than 4 h of light per day; thus this strain will not exploit habitats with short photoperiods, typical of temperate winter or light attenuated depths, even if water temperature warms into the clade HLI tolerance range. MED4 is also excluded from the lowest oxygen habitats of 2.5 µM O~2~. MED4 can, however, grow under OMZ regions with 25 µM O~2~. Genomic analyses (Figure \@ref(fig:ProchlorococcusEnzymeKms))  and transcriptional analyses [@bergResponsesPsbAHli2011]  suggest MED4 is excluded from growth below ~ 25 µM O~2~ because it relies upon a ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction balance in the intersystem electron transport chain.  On the other hand, MED4 shows inducible expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018], to counter photoinactivation of PSII under higher PAR and [O~2~] environments. Photoinactivation, does, however, impose an increased cost of growth upon MED4, since growth under red light, to lower the rate of photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017], allows MED4 to achieve faster growth per absorbed photon than growth under blue light. TARA Oceans Project data [@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of *P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m, representing high to low blue light levels, in the Pacific South East Ocean. Delmont and Eren [@delmontLinkingPangenomesMetagenomes2018] did not analyze data from depths beyond the subsurface chlorophyll maximum layer, the layer in the water column where the chlorophyll *a* concentration peaks, nor did they report the [O~2~] at those depths. Our growth findings are also consistent with Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII proteins annotated as MED4, clade HLI, at depths up to 200 meters with O~2~ of 15 µM.

*Prochlorococcus marinus* SS120, a LLII/III clade representative, showed an interactive inhibition of growth by oxygen and cumulative diel PUR, with a higher tolerance for higher cumulative diel PUR under 25 µM O~2~, compared to 250 µM O~2~ (Figure \@ref(fig:PhotoperiodPurFitsPlots)). Thus, SS120 can exploit habitats with O~2~ levels spanning 25 µM O~2~ to 250 µM O~2~, including higher PAR environments within OMZ.  SS120 is likely excluded from the combination of higher [O~2~] and higher PAR by genomic limitations on capacity for DNA repair (Figure \@ref(fig:DNARepairFig)), and possibly by limited capacity for synthesis of reactive oxygen quenchers (Figure  \@ref(fig:ProchlorococcusEnzymeKms)).  Our growth results are supported by Lavin *et al*. [@lavinNovelLineagesProchlorococcus2010] who found evidence of LLII/III
and LLIV ecotypes, using terminal restriction fragment length polymorphism analyses, at depths above 40 m, where light levels are
higher, within OMZ, and by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII protein subunits annotated as derived from SS120 at all depths ranging
from 20 to 200 m and all [O~2~] in an OMZ of the tropical North Pacific Ocean. SS120 grew under photoperiods longer than 4 h and showed increasing growth rate with increasing photoperiods. SS120 has the potential to thrive in deep temperate zones, specifically during the spring, summer, and fall seasons when the duration of daylight exceeds 4 h, if [O~2~] are near surface saturation of about 250 µM. Under lower oxygen levels of 25 µM, SS120 can also potentially exploit a 4 h photoperiod in the blue waveband, and thus has the potential to inhabit a potential warmed, deep, temperate OMZ, during the winter season.

*Prochlorococcus marinus* MIT9313, a LLIV clade representative, shows potential to inhabit future warmer temperate zones year-round, as it grows under a 4 h photoperiod, expected in winter, or at light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to higher light levels and cumulative diel PUR, but only under low oxygen conditions of 25 µM and 2.5 µM (Figure \@ref(fig:MIT9313PAR)), enabling MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313 carries a gene encoding (S)-2-hydroxy-acid oxidase [@cunaneCrystalStructureAnalysis2005],  with a K~M~ for [O~2~] of ~ 250 µM (Figure \@ref(fig:ProchlorococcusEnzymeKms)), which produces H~2~O~2~.  Growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers photoinhibition, resulting from the inactivation of PSII caused by the production of H~2~O~2~. This photoinhibition is compounded by the limited inducible repair for PSII, due to the absence of FtsH 1 and 2  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  MIT9313 shows remarkable ability to thrive under very low [O~2~], potentially allowing it to expand into broader ecological niches. These results are supported by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSI protein subunits annotated as derived from MIT9313 detected at depths \> 120 m, along with PSII subunits at depths from 50 m to 200 m in regions where O~2~ was 15 µM. Bagby and Chisholm [@bagbyResponseProchlorococcusVarying2015] suggest that O~2~ has a protective role in *Prochlorococcus* under lower carbon dioxide environments when carbon fixation is limited. The deep water environments typical for MIT9313 are relatively nutrient rich, and *Prochlorococcus* take up and metabolize various sugars [@gomez-baenaGlucoseUptakeIts2008; @munoz-marinGlucoseUptakeProchlorococcus2017; @munoz-marinProchlorococcusCanUse2013] and amino acids [@zubkovDepthRelatedAmino2004].  In future work we aim to test whether MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~ environments, or whether PSII generation of O~2~ acts as an electron sink for respiration, using ATP for maintenance and to take up nutrients from the surroundings. Partensky *et al*. [@partenskyComparisonPhotosyntheticPerformances2018] found that in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313 all became net O~2~ consumers, suggesting that the low light levels cause the respiratory chain to consume more O~2~ than the photosynthetic electron transport chain generates, thus contributing to maintenance of the low O~2~ environment.

@@

XXX DOI pending from BOREALIS? or DRYAD? XXX

# Acknowledgements  {.unnumbered}
Carlie Barnhill (Mount Allison Student) assisted with code for import of multicultivator growth data files.
Sarah Arthur (Mount Allison Student) assisted with culturing and setting up Multicultivator runs.

>>>>

*Prochlorococcus marinus*, the smallest picocyanobacterium, comprises multiple clades with distinct niches across tropical and sub-tropical oligotrophic ocean regions, including Oxygen Minimum Zones.  Ocean warming may open permissive temperatures in  new, poleward photic regimes, along with expanded Oxygen Minimum Zones.  We used ocean protein data [CITATION] to help guide testing of *Prochlorococcus marinus* growth across a matrix of peak irradiances, photoperiods, spectral bands and dissolved oxygen.  MED4 from Clade HLI requires greater than 4 h photoperiod, grows at 25 µmol O~2~ L^-1^ and above, and exploits high cumulative diel photon doses, yet shows accelerated growth when the cost of photoinactivation is lowered under red, vs. blue, actinic light.  MED4 relies upon an alternative oxidase to balance electron transport, which may exclude it from growth under our lowest 2.5 µmol O~2~ L^-1^ condition.  SS120 from Clade LLIII is restricted to low light under full 250 µmol O~2~ L^-1^ , shows expanded light exploitation under  25 µmol O~2~ L^-1^, but is excluded from growth under  2.5 µmol O~2~ L^-1^.  Intermediate oxygen suppresses the cost of PSII photoinactivation, and enzymatic production of H~2~O~2~ in SS120, which has limited genomic capacity for PSII and DNA repair.  MIT9313 from Clade LLIV is restricted to low blue irradiance under 250 µmol O~2~ L^-1^, but exploits much higher irradiance under red light, or under lower [O~2~], conditions which slow photoinactivation of PSII and production of reactive oxygen species.

@@

We analyzed growth rates for  *P. marinus* clade HLI found near the ocean surface; clade LLII/III found deep in the water column; and clade LLIV also found in deep oceans, including OMZ, under a matrix of spectral wavebands, irradiances, photoperiods and oxygen concentrations approximating present day and hypothetical future niches.

*Prochlorococcus marinus* MED4 requires more than 4 h of light per day; thus this strain will not exploit habitats typical of temperate winter or light attenuated depths, even if water temperatures warm into the clade HLI tolerance range. MED4 is also excluded from the lowest oxygen habitats of 2.5 µM O~2~, but can, grow under OMZ regions with 25 µM O~2~. Genomic (Figure \@ref(fig:ProchlorococcusEnzymeKms)) and transcriptional analyses [@bergResponsesPsbAHli2011]  suggest MED4 is excluded from growth below ~ 25 µM O~2~ because it relies upon a ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction balance in the intersystem electron transport chain, with a K~M~ for  [O~2~] of ~25 µM O~2~.  On the other hand, MED4 shows inducible expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018], to counter photoinactivation of PSII under higher PAR and [O~2~] environments. Photoinactivation, does, however, impose an increased cost of growth upon MED4, since growth under red light, to lower photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017], allows MED4 to achieve faster growth per absorbed photon than growth under blue light. TARA Oceans Project data [@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of *P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m, representing high to low blue light levels, in the Pacific South East Ocean. Delmont and Eren [@delmontLinkingPangenomesMetagenomes2018] did not analyze data from depths beyond the subsurface chlorophyll maximum layer, nor did they report[O~2~] at depth. Our growth findings are  consistent with Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII proteins annotated as MED4, clade HLI, at depths up to 200 meters, with O~2~ of ~15 µM.

*Prochlorococcus marinus* SS120, a LLII/III clade representative, showed an interactive inhibition of growth by oxygen and cumulative diel PUR, with a higher tolerance for higher cumulative diel PUR under 25 µM O~2~, compared to 250 µM O~2~ (Figure \@ref(fig:PhotoperiodPurFitsPlots)). Thus, SS120 can exploit higher PAR environments, within OMZ.  SS120 is likely excluded from the combination of higher [O~2~] and higher PAR by genomic limitations on capacity for DNA repair (Figure \@ref(fig:DNARepairFig)), and possibly by limited capacity for synthesis of reactive oxygen quenchers (Figure  \@ref(fig:ProchlorococcusEnzymeKms)).  Our growth results are supported by Lavin *et al*. [@lavinNovelLineagesProchlorococcus2010] who found evidence of LLII/III and LLIV ecotypes, using terminal restriction fragment length polymorphism analyses, at depths above 40 m, where light levels are higher, within OMZ, and by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII protein subunits annotated as derived from SS120 at all depths ranging from 20 to 200 m and all [O~2~] in an OMZ of the tropical North Pacific Ocean. SS120 grew under photoperiods longer than 4 h and showed increasing growth rate with increasing photoperiods, and so has the potential to thrive in deep temperate zones, specifically during the spring, summer, and fall seasons when the duration of daylight exceeds 4 h, if [O~2~] are near surface saturation of about 250 µM. Under lower oxygen levels of 25 µM, SS120 can also potentially exploit a 4 h photoperiod in the blue waveband, and thus has the potential to inhabit a potential warmed, deep, temperate OMZ, during the winter season.

*Prochlorococcus marinus* MIT9313, a LLIV clade representative, shows potential to inhabit future warmer temperate zones year-round, as it grows under a 4 h photoperiod, expected in winter, or at light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to higher light levels and cumulative diel PUR, but only under low oxygen conditions of 25 µM and 2.5 µM (Figure \@ref(fig:MIT9313PAR)), enabling MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313 carries a gene encoding (S)-2-hydroxy-acid oxidase [@cunaneCrystalStructureAnalysis2005],  with a K~M~ for [O~2~] of ~ 250 µM (Figure \@ref(fig:ProchlorococcusEnzymeKms)), which produces H~2~O~2~.  Growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers photoinhibition, resulting in part from the inactivation of PSII caused by the production of H~2~O~2~. This photoinhibition is compounded by limited inducible repair for PSII, due to the absence of FtsH 1 and 2  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  MIT9313 shows remarkable ability to thrive under very low [O~2~], potentially allowing it to expand into broader ecological niches. These results are supported by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSI protein subunits annotated as derived from MIT9313 detected at depths \> 120 m, along with PSII subunits at depths from 50 m to 200 m in regions where O~2~ was 15 µM. Bagby and Chisholm [@bagbyResponseProchlorococcusVarying2015] suggest that O~2~ has a protective role in *Prochlorococcus* under lower carbon dioxide environments when carbon fixation is limited. The deep water environments typical for MIT9313 are relatively nutrient rich, and *Prochlorococcus* take up and metabolize various sugars [@gomez-baenaGlucoseUptakeIts2008; @munoz-marinGlucoseUptakeProchlorococcus2017; @munoz-marinProchlorococcusCanUse2013] and amino acids [@zubkovDepthRelatedAmino2004].  In future work we aim to test whether MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~ environments, or whether PSII generation of O~2~ acts as an electron sink for respiration, using ATP for maintenance and to take up nutrients from the surroundings. Partensky *et al*. [@partenskyComparisonPhotosyntheticPerformances2018] indeed found that in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313 all became net O~2~ consumers, suggesting that the low light levels cause the respiratory chain to consume more O~2~ than the photosynthetic electron transport chain generates, thus contributing to maintenance of the low O~2~ environment.

@@

# Acknowledgements  {.unnumbered}
Carlie Barnhill (Mount Allison Student) assisted with code for import of multicultivator growth data files.
Sarah Arthur & Daytona McMackin (Mount Allison Students) assisted with culturing and setting up Multicultivator runs.
